### 2020-01-16 manuscript/manuscript.rmd
update manuscript

## Abstract
Phylogenomic studies based on hundreds to thousand of genes have greatly improved our understanding of evolutionary relationship among major animal lineages. However, a considerable debate still remains regarding to the animal origin, with Ctenophora-sister and Porifera-sister emerging as the two primary hypotheses, comprising one of the most difficult node to resolve in animal tree of life. Here, we systematically explore the mechanisms that potentially cause the incongruent results by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses in these data in a consistent manner to characterize the differences between studies. We find the gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices. More Importantly, We find Porifera-sister can only be recovered with more constraints, including using site-heterogeneous CAT model with inclusive of the most closely related outgroups in several matrices, whereas Ctenophora-sister is recovered in all other cases. We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences of distribution of phylogenetic signal between models and outgroup choice (especially CAT models). Last, we show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses. Our results provide a comprehensive overview of current understanding of the first diverging animal branch and why different studies yield different answers. Rather than embracing one hypothesis over another, we hope our efforts provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

@@

Phylogenomic analyses with genome-scale data have the potential to revolutionize our understanding of the tree of life. However, over the past decade, there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister (comb jelly) and Porifera-sister (sponges) (Fig XXOverview) emerging as the two primary hypotheses. Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested. In contrast to the lack of debate about the position of Porifera, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. An understanding of the first diverging animal lineage is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for the evolution of complex structures such as cell types, nervous systems, and developmental patterns. However, a wide consensus of animl origin has not been reached despite multiple efforts.

The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative. Since then many more studies have been published and extensive reanalyzed, some supporting Ctenophora-sister, some Porifera-sister, and some neither (Table 1). As it has become clear that this is a very difficult phylogenetic challenge, and the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem. Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support. Here we synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, and reanalyze these data using standardized methods, and perform new analyses to characterize differences between studies. We hope that this effort provides an integrative overview of the challenge and provides direction for future studies. We also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop methods to address difficult phylogenetic problems.

@@

Models of molecular evolution have several components that each consider different aspects of the evolutionary process. The models that have been used to model protein evolution in studies of the deep phylogeny have largely differed according to three components: the exchangeability matrix $E$, the rate of evolution, and the state equilibrium frequencies $\Pi$. A detailed description of each component is in Supplementary text. Briefly, (1) the exchangeability matrix $E$ describes the rate at which one amino acid changes to another. Exchangeability matrices have been used in the studies under consideration here include: Poisson (or F81), WAG, LG, GTR, as they have been used regularly in phylogenomic analyse; (2) While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled, such as site homogeneous rates and Gamma-rate heterogeneity; (3) the vector of equilibrium frequencies $\Pi$ describes the stationary frequency of amino acids, such as empirical site-homogeneous, estimated site-homogeneous, CAT site-heterogeneous [@Lartillot:2004dq], and C10-C60 models (C models). (4) data-recoding method recode 20 amino acids are recoded into six groups based on function to account for both compositional heterogeneity and substitution saturation. Several recoding strategies have been proposed, including Dayhoff 6-state recoding, S&R 6-state recoding, KGB 6-state recoding based on exchangeability matrix of models [@feuda2017improved].

Models can be assembled by selecting different options for all these different components. The models that are applied in practice area heavily influenced by engineering and computational costs, as well as convention. For example, on the questions considered here, Poisson and GTR exchangeability matrices have only been used in combination with CAT site-heterogeneous models of equilibrium frequency. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to variation across sites and WAG a particular exchangeability matrix. CAT is generally shorthand for F81+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. One could, though, run a WAG+CAT model. To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT) or rate (*e.g.*, Gamma). If there are no modifiers, then it is implied that site homogeneous models are used.

@@

Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect the phylogenetic inference of animal evolution (**Fig XXOverview.C**). Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated, including Choanomalia (Metazoa plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi). It has been suggested the inclusion of more distant related (e.g. Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous CAT models [@pisani2015genomic]. In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention than outgroup sampling. This may be because results have tended to be more sensitive to outgroup sampling.

@@

Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position that needs to be explored. However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages (Fig XXTaxon_composition). These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting]). Importantly, several earlier studies only include less than five ctenophore species (Fig XXTaxon_composition). The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies a lot. In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of the earliest diverging animal lineage. Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other matrices (Table1, Fi. XXTaxon_composition).

```{r taxon_rectangles }

@@

We first used the Metazoa BUSCO genes [@waterhouse2017busco0] to evaluate gene composition from all data matrices and the result is unexpected. The animal BUSCO dataset contains 978 single-copy OGs (identified from 64 animal taxa) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of genome or transcriptomic assemblies [@waterhouse2017busco]. Moreover, the BUSCO genes also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics]). Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices. However, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Fig. XXBUSCO_annotations).

@@

Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig. XXsupport_published_analyses and Table XXperviouly_published_analyses). Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 supports Ctenophora-sister (Table 1). Moreover, three of them are reanalyzing studies without generating any new data matrices. Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used. The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT+G or GTR+CAT+G) models. It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes. The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT+G and recoding methods with GTR+CAT+G models in several matrices. Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only Choanofalgellates are used as outgroups (Fig. support_published_analyses). Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups, whereas Ctenophora-sister is supported in all other cases.

```{r support_published_analyses}

analyses_published %>%
	ggplot( aes(x=model_combined, y=clade, col=result, shape=inference) ) +
		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
		scale_colour_manual(values = result_colors) +
		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
   xlab("model complexity")
```

@@

```{r support_new_analyses}

analyses_new %>%
	ggplot( aes(x=model_summary, y=clade, col=result, shape=inference) ) +
		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
		scale_colour_manual(values = result_colors) +
		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))
```

One of the challenges of interpreting support for the placement of the animal root across studies is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways. Here we first reanalyze the primary matrices from each study under consistent conditions with IQtree [@nguyen2014iq] under a panel of evolutionary models. We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework [@zhou2017evaluating]. Importantly, we also include a CAT-like, site-heterogeneous C10 - C60 (C) models [@si2008empirical] that are implemented in both ML and Bayesian inference (BI) framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.

Here we reanalyzed 17 from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table 1). We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig XXOverview). We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60+F+G models are used in 3 large matrices). We found in all cases, WAG/LG+C60+G models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree. It should be noted that we didn't include GTR+C models in model testing since it is computational too expensive. The GTR+C models requires to estimate up to over 10,000 (e.g. GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous + C models. We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG+G, GTR+G and Poisson+C60+F+G. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups. Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups.

@@

### Comparison of iqtree and phylobayes results

Site heterogeneity in equilibrium frequency has been a major concern in tests of Ctenophora-sister and Porifera-sister. This has been addressed with CAT models. IQtree provides a new family of C models that also address site heterogeneity. Given the extensive computational cost and concerns about overparameterization of CAT models [@whelan2016let], we compared iqtree C results to CAT results for a subset of matrices to see if they give consistent results. This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses. For computational efficiency, we first only applied Poisson+CAT+G (Poisson+CAT) models in every matrix with Choanoflagellata as outgroups. It should be noted that most Phylobayes runs were converged, although several large matrices have not reached convergence after at least a month's computational time.

@@

The distribution of phylogenetic signal has never been quantified in site-heterogeneous models and compared with site-homogeneous models. To further explore the effect of model selection and outgroup choices in animal origin, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig. 1A); T2: Porifera-sister (Fig. 1B)) to three key data matrices with different outgroup choice and models in both ML and BI framework (Fig XXPhylogenetic signal). By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS\) or site (\Delta$SLS\) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes). Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in site-homogeneous models. This finding is largely consistent with the past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].

In contrast, the phylogenetic signal of site-heterogenous models changed dramatically comopared to the site-homogenous models. Although Ctenophora-sister also had the higher proportions of supporting genes with C models, we found that the phylogenetic signal decreases in many genes using C models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig. XXphylogenetic_signal B). It is not clear why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing in IQtree, but our results indicated the C models lack of phylogenetic signal compared to site-homogeneous models in the node of the root position of animal phylogeny. In contrast to the C models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Fig. XXphylogenetic_signalA, C). We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models. The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.

@@

```{r phylogenetic_signal, fig.width=8, fig.height=12}
#au_tests %>%
#  unite(Matrix, c(manuscript, matrix, model)) %>%
#  group_by(Matrix, result) %>%
#  count() %>%
#  group_by(Matrix) %>%
#  mutate(percent = prop.table(n)) %>%

@@

x1  = factor(Philippe2009$Matrix, levels=c("Philippe2009_WAG", "Philippe2009_Choanozoa_WAG", "Philippe2009_LG", "Philippe2009_Choanozoa_LG", "Philippe2009_LG(phylobayes)", "Philippe2009_Choanozoa_LG(phylobayes)", "Philippe2009_C60", "Philippe2009_Choanozoa_C60", "Philippe2009_F81", "Philippe2009_Choanozoa_F81"))
y1=factor(Philippe2009$result, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))

@@

y2=factor(Ryan2013_est$result, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))

x2  = factor(Ryan2013_est$Matrix, levels=c("Ryan2013_est_WAG", "Ryan2013_Choanozoa_WAG", "Ryan2013_est_LG", "Ryan2013_Choanozoa_LG",  "Ryan2013_est_C60", "Ryan2013_Choanozoa_C60"))

Whelan2017 = filter(au_tests,
                   grepl('Whelan', Matrix))
y3=factor(Whelan2017$result, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))

x3  = factor(Whelan2017$Matrix, levels=c("Whelan2017_full_WAG", "Whelan2017_Choanozoa_WAG", "Whelan2017_full_LG", "Whelan2017_Choanozoa_LG", "Whelan2017_full_LG(phylobayes)", "Whelan2017_Choanozoa_LG(phylobayes)", "Whelan2017_full_C60", "Whelan2017_Choanozoa_C60", "Whelan2017_full_F81", "Whelan2017_Choanozoa_F81"))

>>>>

## Abstract
Phylogenomic studies based on hundreds to thousands of genes have greatly improved our understanding of evolutionary relationships among major animal lineages. However, a considerable debate still remains regarding the animal origin, with Ctenophora-sister and Porifera-sister emerging as the two primary hypotheses, comprising one of the most difficult to resolve nodes in animal tree of life. Here, we systematically explore the mechanisms that potentially cause the incongruent results by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses with these data in a consistent manner to characterize the differences between studies. We find that gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices. More importantly, we find Porifera-sister can only be recovered with more constraints, including using site-heterogeneous CAT model inclusive of the most closely related outgroups in several matrices, whereas Ctenophora-sister is recovered in all other cases. We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences in the distribution of phylogenetic signal between models and outgroup choice (especially CAT models). Last, we show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses. Our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers. Rather than embracing one hypothesis over another, we hope our efforts provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

@@

Phylogenomic analyses with genome-scale data have the potential to revolutionize our understanding of the tree of life. However, over the past decade, there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister (comb jelly) and Porifera-sister (sponges) (Fig XXOverview) emerging as the two primary hypotheses. Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested. By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. An understanding of the first diverging animal lineage is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for the evolution of complex structures such as cell types, nervous systems, and developmental patterns. However, a wide consensus of animl origin has not been reached despite multiple efforts.

The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative. Since then many more studies have been published and extensive reanalyzed, some supporting Ctenophora-sister, some Porifera-sister, and some neither (Table 1). It has become clear that this is a very difficult phylogenetic challenge, and as the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem. Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support. Here we synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, and reanalyze these data using standardized methods, and perform new analyses to characterize differences between studies. We hope that this effort provides an integrative overview of the challenge and provides direction for future studies. We also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop methods to address difficult phylogenetic problems.

@@

Models of molecular evolution have several components that each consider different aspects of the evolutionary process. The models that have been used to model protein evolution in studies of the deep phylogeny have largely differed according to three components: the exchangeability matrix $E$, the rate of evolution, and the state equilibrium frequencies $\Pi$. A detailed description of each component is in Supplementary text. Briefly, (1) the exchangeability matrix $E$ describes the rate at which one amino acid changes to another. Exchangeability matrices have been used in the studies under consideration here include: Poisson (or F81), WAG, LG, GTR, as they have been used regularly in phylogenomic analyses; (2) While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled, such as site homogeneous rates and Gamma-rate heterogeneity; (3) the vector of equilibrium frequencies $\Pi$ describes the stationary frequency of amino acids, such as empirical site-homogeneous, estimated site-homogeneous, CAT site-heterogeneous [@Lartillot:2004dq], and C10-C60 models (C models). (4) data-recoding method recode 20 amino acids are recoded into six groups based on function to account for both compositional heterogeneity and substitution saturation. Several recoding strategies have been proposed, including Dayhoff 6-state recoding, S&R 6-state recoding, KGB 6-state recoding based on exchangeability matrix of models [@feuda2017improved].

Models can be assembled by selecting different options for all these different components. The models that are applied in practice area heavily influenced by engineering and computational costs, as well as convention. For example, on the questions considered here, Poisson and GTR exchangeability matrices have only been used in combination with CAT site-heterogeneous models of equilibrium frequency. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to variation across sites and WAG a particular exchangeability matrix. CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. One could, though, run a WAG+CAT model, although the computation time is much longer than Poisson+CAT model. To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT). If there are no modifiers, then it is implied that site homogeneous models are used. Moreover, the Gamma-rate heterogeinity is used in almost every analysis.

@@

Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect the phylogenetic inference of animal evolution (Fig XXOverview.C). Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated, including Choanomalia (Metazoa plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi). It has been suggested the inclusion of more distant related (e.g. Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous CAT models [@pisani2015genomic]. In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention than outgroup sampling. This may be because results have tended to be more sensitive to outgroup sampling.

@@

Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position is poorly understood. However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages (Fig XXTaxon_composition). These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting]). Importantly, several earlier studies only include less than five ctenophore species (Fig XXTaxon_composition). The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies a lot. In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of the earliest diverging animal lineage. Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other matrices (Table1, Fi. XXTaxon_composition).

```{r taxon_rectangles, fig.height = 12, fig.width = 20 }

@@

We first used the Metazoa BUSCO genes [@waterhouse2017busco0] to evaluate gene composition from all data matrices and the result is unexpected. The animal BUSCO dataset contains 978 single-copy OGs (identified from 64 animal taxa) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of genome or transcriptomic assemblies [@waterhouse2017busco]. Moreover, the BUSCO genes also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics]). Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices. However, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Fig  XXBUSCO_annotations).

@@

Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig  XXsupport_published_analyses and Table XXperviouly_published_analyses). Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Table 1). Moreover, three studies are secondary studies without any new data matrices. Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used. The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models. It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes. The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models in several matrices. Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only Choanofalgellates are used as outgroups (Fig support_published_analyses). Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups, whereas Ctenophora-sister is supported in all other cases.

```{r support_published_analyses}
analyses_published=read_tsv("../data_processed/tables/previously_published_analyses.tsv")
x1 = factor(analyses_published$model_combined, levels=c("WAG", "LG", "GTR", "data partitioning", "Recoding + GTR", "Recoding + GTR + CAT", "Poisson + CAT", "GTR + CAT"))
analyses_published %>%
	ggplot( aes(x=x1, y=clade, col=final, shape=inference) ) +
		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
		scale_colour_manual(values = result_colors) +
		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
   xlab("Model complexity") + ylab("Clade")
```

@@

```{r support_new_analyses}
analyses_new=analyses_new %>%
mutate(model=recode(model,
                    `GTR20`="GTR",
                    `CAT+F81`="Poisson+CAT"))

x1 = factor(analyses_new$model, levels=c("WAG", "GTR", "Poisson+C60", "WAG+C60", "LG+C60", "Poisson+CAT"))
analyses_new %>%
	ggplot( aes(x=x1, y=clade, col=result, shape=inference) ) +
		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
		scale_colour_manual(values = result_colors) +
		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
       xlab("Model complexity") + ylab("Clade")
```

One of the challenges of interpreting support for the placement of the animal root is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways. Here we first reanalyze the primary matrices from each study under consistent conditions with IQtree [@nguyen2014iq] under a panel of evolutionary models. We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework [@zhou2017evaluating]. Importantly, we also include a CAT-like, site-heterogeneous C10 - C60 (C) models [@si2008empirical] that are implemented in both ML and Bayesian inference (BI) framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.

Here we reanalyzed 17 matrices from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table 1). We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig XXOverview). We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60 models are used in 3 large matrices). We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree. It should be noted that we didn't include GTR+C models in model testing since it is computational too expensive. The GTR+C models requires to estimate up to over 10,000 (e.g. GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous + C models. We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60+F. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups. Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups.

@@

### Comparison of IQtree and PhyloBayes results

Site heterogeneity in equilibrium frequency has been a major concern in tests of Ctenophora-sister and Porifera-sister. This has been addressed with CAT models. IQtree provides a new family of C models that also address site heterogeneity. Given the extensive computational cost and concerns about overparameterization of CAT models [@whelan2016let], we compared iqtree C results to CAT results for a subset of matrices to see if they give consistent results. This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses. For computational efficiency, we first only applied Poisson+CAT (Poisson+CAT) models in every matrix with Choanoflagellata as outgroups. It should be noted that most Phylobayes runs were converged, although several large matrices have not reached convergence after at least a month's computational time.

@@

The distribution of phylogenetic signal has never been quantified in site-heterogeneous models and compared with site-homogeneous models. To further explore the effect of model selection and outgroup choices in animal origin, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig. 1A); T2: Porifera-sister (Fig. 1B)) to three key data matrices with different outgroup choice and models in both ML and BI framework (Fig XXPhylogenetic signal). By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS\) or site (\Delta$SLS\) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes). Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in sifte-homogeneous models. This finding is largely consistent with the past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].

In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models. Although Ctenophora-sister also had the higher proportions of supporting genes with C models, we found that the phylogenetic signal decreases in many genes using C models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig. XXphylogenetic_signal B; tableXXdistribution). It is not clear why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing in IQtree, but our results indicated the C models lack of phylogenetic signal compared to site-homogeneous models in the node of the root position of animal phylogeny. In contrast to the C models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Fig. XXphylogenetic_signalA, C). We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (tableXXdistribution). The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.

@@

```{r phylogenetic_signal, fig.width=6, fig.height=12}
#au_tests %>%
#  unite(Matrix, c(manuscript, matrix, model)) %>%
#  group_by(Matrix, result) %>%
#  count() %>%
#  group_by(Matrix) %>%
#  mutate(percent = prop.table(n)) %>%

@@

x1  = factor(Philippe2009$Matrix, levels=c("Philippe2009_WAG", "Philippe2009_Choanimalia_WAG", "Philippe2009_LG", "Philippe2009_Choanimalia_LG", "Philippe2009_LG (PhyloBayes)", "Philippe2009_Choanimalia_LG (PhyloBayes)", "Philippe2009_WAG+C60", "Philippe2009_Choanimalia_WAG+C60", "Philippe2009_Poisson+CAT (PhyloBayes)", "Philippe2009_Choanimalia_Poisson+CAT (PhyloBayes)"))
y1=factor(Philippe2009$Catogary, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))

@@

y2=factor(Ryan2013_est$Catogary, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))

x2  = factor(Ryan2013_est$Matrix, levels=c("Ryan2013_est_WAG", "Ryan2013_Choanimalia_WAG", "Ryan2013_est_LG", "Ryan2013_Choanimalia_LG",  "Ryan2013_est_WAG+C60", "Ryan2013_Choanimalia_WAG+C60"))

Whelan2017 = filter(au_tests,
                   grepl('Whelan', Matrix))
y3=factor(Whelan2017$Catogary, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))
